---
title: "01_plotting_ggplot"
author: "JR"
date: "2024-07-29"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)

```
# Youtube video for this class
https://youtu.be/9DUHEOP7LuA

# Goal: Learn how to plot our data using GGPLOT and the coding logic in ggplot !
One of the best parts of R is the plotting abilities. You are, by analogy designing your own figures with code. You can run many stastical analyses and plot the outputs in the code to make a figure. Here we will focus on GGPLOT(GGPLOT2). The GG says it all:

"Grammar of Graphics" -- it truly is a very pure and flexible way of plotting data for figures!

There are 8 possible layers of information -- each building on the data layer that you are importing. This looks really confusing but we will slowly build up from the basics (mapping, stats, geom)

Plotting layers of R:

1) Data -- data being plotted from object indexes etc.

2) MAPPING -- mapping data (aes)

3) Statistics -- statical analyses on mapped data

4) scales -- how big small you want things

5) GEOMETRIES -- what type of plots you are making (geom)

6) Facets -- making multiple representations of plots

7) Coordinates -- specifing placement of image aspects

8) THEME -- a set version of colors, background. Mostly so you don't have to add many of the layers above everytime!

Ok let's walk through some simple examples to more difficult figures using these layers!

Layer 1 : DATA (data types matter so good to label object accrodingly)

# loading our results from lessons/06_Differential_expression_analyses/results/TPM_results
# and lessons/06_Differential_expression_analyses/results/counts_results_DESEQ
```{r loading our RNAseq analysis data TPM & COUNTS}

# load TPM RESULTS: 
load("/scratch/Shares/rinnclass/MASTER_CLASS/lessons/06_Differential_expression_analyses/results/TPM_results/TPM_results.Rdata")

# laod DESEQ2 results
load("/scratch/Shares/rinnclass/MASTER_CLASS/lessons/06_Differential_expression_analyses/results/counts_results_DESEQ/DESEQ_results.rdata")

```

# (1-2) GGPLOT DATA & MAPPING LAYERS: basic example
Now let's start plotting in GGplot!
```{r ggplot basics}

# ggplot first wants the data parameter (layer 1), then an 'aesthetics' layer (aes)
# aes() layer needs everything you want to do with the data. The + adds the next layer (layer 2).
?ggplot

# let's make this simple plot first: 
ggplot(filtered_res_df, aes(x = log2FoldChange, y = padj)) +
  geom_point()
# Voila we have a figure and a result! Most sig changes are in the .
# Let's double check this result by changing the x-axis

ggplot(filtered_res_df, aes(x = log2FoldChange, y = padj)) +
  geom_point() +
  xlim(-10, 10)

# Futher zoom in
ggplot(filtered_res_df, aes(x = log2FoldChange, y = padj)) +
  geom_point() +
  xlim(-5, 5)

# Yup most of our data has lf2c less than absolute value of 2.5-ish
# Note you can see how many genes were removed by each x-axim limit change in console
# PS this is a "volcano plot
```


# for fun you can change a lot of shapes and colors in plots: changing shapes/colore in plot
Now let's change the shape of the points outside the aes() layer and in the geom() of the mapping layer :)
```{R changing shapes in plot}
             
ggplot(filtered_res_df, aes(x = log2FoldChange, y = padj)) +
  geom_point(shape = 'square', color = 'purple') +
  xlim(-5, 5) 
  
# Here are some good resources for other shapes and color vignettes:
# http://www.sthda.com/english/wiki/ggplot2-point-shapes

```

# coloring by data in aes() layers
We colored the plot above in the geom layer, but what if we wanted to color by "is_super_sig"?
Since that depends on the data we need to do that in the aes() layer!
We can even index and run functions in aes()
```{R # coloring by data in aes() layer}


ggplot(filtered_res_df,
       aes(x = log2FoldChange, y = padj, color = result_name)) +
  geom_point()

# Nice now we can see that we mapped our data in the aes() layer to be by timepoint of sigifcance.
# Let's just color the earliest time points going deeper into the mapping layer aes() layer

ggplot(filtered_res_df, aes(x = log2FoldChange, 
                            y = padj, 
                            color = result_name == "time_point_96_vs_0")) +
  geom_point()

```
# RESULT: interesting we see that the latest time points have the largest fc


# facet in ggplot to make multiple figures
Facet let's you pick a variable and make all the same figures for that variable.
Let's try it out with each time point
```{R facet by timepoint}

ggplot(filtered_res_df, aes(x = log2FoldChange, 
                            y = padj)) +
         facet_wrap(result_name ~ .) +
  geom_point() 

# Nice we have 4 plots one for each time point !
  
```
# RESULT: we see that early and late time point have the latest changes
# RESULT: We see that the largest fold change occurs early and then more show up late



# Geom layer : histogram
Now let's take a look at other types of plots in the geom layer.
geom_histogram() is very useful in a lot of analyses. Let's start there.

```{r Geom_histogram}

# frequencty of peak numbers across DBPs

ggplot(filtered_res_df, aes(x = log2FoldChange)) +
  geom_histogram()

?geom_histogram()
# Result: peak number is represented by right gaussian.
# let's add more bins

ggplot(filtered_res_df, aes(x = log2FoldChange)) +
  geom_histogram(bins = 300) +
   xlim(-5, 5) 

# Let's chage the color of the bars in histogram by number of peaks.
# again we need to do this inside aes() since the data needs to read.

ggplot(filtered_res_df, aes(x = log2FoldChange, fill = result_name)) +
  geom_histogram(bins = 300) +
   xlim(-5, 5) 

# this is hard to see the four distributions.
# let's fix this by adding a 'position' parameter in the geom layer.

ggplot(filtered_res_df, aes(x = log2FoldChange, fill = result_name)) +
   geom_histogram(bins = 300, position = "dodge") +
   xlim(-2, 2) 

# so now we can see the individual data points separated!

```

# Geom layer : density plot
DENISTY Plots are also very useful to see where the density of data arises.
```{R geom_density()}

ggplot(filtered_res_df, aes(x = log2FoldChange , fill = result_name)) +
   geom_density()

# but again its hard to see let's fix that with shading or alpha in
# geom layer

ggplot(filtered_res_df, aes(x = log2FoldChange, fill = result_name)) +
   geom_density(alpha = 0.1)

# let's change color of the line now (inside aes()) and also use a function inside aes()

ggplot(filtered_res_df,
       aes(
         x = log2FoldChange,
         fill = padj < 0.01,
         color = result_name
       )) +
  geom_density(alpha = 0.3) +
  xlim(-2, 2)

# so now line and fill are same color -- let's see:

ggplot(filtered_res_df,
       aes(x = log2FoldChange, fill = result_name, color = result_name)) +
  geom_density(alpha = 0.0)+
  xlim(-2,2)

```
# RESULT: the 12 hour time point has the largest amounts of changes
# RESULT: there is more upregulation at 12 hours than other timepoints

AWESOME, we now can plot our data and visualize our results. Yet there is much
more to learn in GGPLOT so we will pick up here on the next lesson !

