---
title: "02_GGPLOT_2"
author: "JR"
date: "2024-08-03"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(magrittr)
# install.packages("broom")
library(broom)

```

# Load data
```{r loading our RNAseq analysis data TPM & COUNTS}

# load TPM RESULTS: 
load("/scratch/Shares/rinnclass/MASTER_CLASS/lessons/06_Differential_expression_analyses/results/TPM_results/TPM_results.Rdata")

# laod DESEQ2 results
load("/scratch/Shares/rinnclass/MASTER_CLASS/lessons/06_Differential_expression_analyses/results/counts_results_DESEQ/DESEQ_results.rdata")

```

# Alright in the last class we eneded on the GEOM layer in GGPLOT
# To review we are up to here:
1) Data -- data being plotted from object indexes etc.
2) MAPPING -- mapping data (aes)
3) Statistics -- statical analyses on mapped data
4) scales -- how big small you want things
5) GEOMETRIES -- what type of plots you are making (geom)

# We explored density and histogram Geometries now let's look at some more !


# Multiple Geom layers 
The geom layer has a lot of options AND you can layer them!
Let's take out x,y plots above and map them into a 2D density plot
```{R geom_point() & geom_density_2d}

ggplot(filtered_res_df, aes(x = baseMean, 
                 y = padj)) +
  geom_point()+
  xlim(0,1000) +
  geom_density_2d() 

# This now shows us where the density of our data is !

# Now let's compare log2fc and lfcse
ggplot(filtered_res_df, aes(x = log2FoldChange, 
                 y = lfcSE)) +
  geom_point()+
  xlim(-5,5) +
  geom_density_2d() 

# what would happen if we change the order?

ggplot(filtered_res_df, aes(x = log2FoldChange, 
                 y = lfcSE)) +
  geom_density_2d() +
  xlim(-5,5) +
  geom_point()
  

# it's hard to see but in this case the contour is on the bottom
#  -- subtle but important

```
# RESULT: we see a linear relationship between l2fc and lfcse


# trend lines
Let's add even more geom_*() layers -- starting with a 'trend line'
```{r geom_smooth, geome_abline}

# We might suspect that padj and pvalue have a linear realtionship.
# Let's take a look:

ggplot(filtered_res_df, aes(x = padj, 
                 y = pvalue)) +
  geom_point()

# Yup seems like it - here is where trend lines come in handy !

?geom_abline
# This will make a line with a given slope and intercept
?geom_smooth
# This shows variance and deviations of the trend lines
# let's add them both !

ggplot(filtered_res_df, aes(x = padj, 
                 y = pvalue)) +
  geom_point()+
  geom_abline(slope = 1, intercept = 0) +
  geom_smooth(method = "lm")

# Nice we see that the linear model fits well (blue)
# Try changing the slope to identify the relationship !?

```
# RESULT: we observe a linear trend between pvalue and padj

# Now let's retreive the values in these linear models
Above we just plotted the lines on top of data. But now let's 
get the actual values of the linear regression fits! 

We just need to pick to variables and find the regression between them
To do this we use the function "lm" and assign values to it 
```{r linear model data}

lm_fit <- lm(filtered_res_df$padj ~ filtered_res_df$pvalue)

# IMPORTANT: Summary and table are super handy functions in R
summary(lm_fit)
# Nice summary gave us the results of the linear fit we could add to the plot

summary(filtered_res_df$baseMean)
# WOW super handy - let's look more :) !
summary(filtered_res_df$padj)
```

# Digging deeper on trend lines and plotting values !
```{r obtaining and plotting trend lines with broom }

# ORDER MATTERS: 
# Here is what we did above:
lm_fit <- lm(filtered_res_df$padj ~ filtered_res_df$pvalue)
summary(lm_fit)

# Now let's swap the values and see what happens 
lm_fit_2 <- lm(filtered_res_df$pvalue ~ filtered_res_df$padj)
summary(lm_fit_2)

# Ok make sure to think through the order !

# let's convert the lm_fit data to a data_frame using tidy()
# it's in the library broom:
library(broom)
lm_df <- tidy(lm_fit)
lm_df
# so much easier to read than summary :)

model <- lm(filtered_res_df$padj ~ filtered_res_df$pvalue)
summary(model)
# plotting with the actual lm fit:

ggplot(filtered_res_df, aes(x = padj, y = pvalue)) +
  geom_point() +
  geom_abline(slope = 1/(lm_df$estimate[[2]]))

# See how this changes if order is changed (the first one is Y and then X)
# Thus why we have 1/lm_df$estimate

```
# RESULT: Pvalue is about 3.5 times lower than padj.


# More GEOM(_) layers : Bar plots
Let's take a quick look at bar plots 
we provide the x-axis and ggplot counts for Y.
We will also use this opportunity to look at level 3) STAT layer
```{r geom_bar()}
# let's plot the number of each type of DBD in our set of DBPs 

ggplot(filtered_res_df, aes(x = result_name)) + 
  geom_bar()
# Nice, we see that GGPLOT automatically made the y-axis the values in result_name
# We observe that the most changes occured at 48 hours

# If the we want to provide the Y-axis for a specific DBP
# we need to apply stat = identity in the geom_bar()
# basically this allows you to command the y-axis

ggplot(filtered_res_df, aes(
      x = result_name,
      y = baseMean)) +
        geom_bar(stat = "identity")

# Interesting what are those y-values :) ?
# Guessing they are the sum of the baseMean in each time_point
# Let's test and find out :
 
sum_basemean <- filtered_res_df %>%
  filter(result_name == "time_point_12_vs_0") %>%
  summarise(sum_basemean = sum(baseMean))
# YUP, so because we set Y-axis to basemean it took the sum of all basemeans !
```
# RESULT: base mean is also highest at 48 hours
# RESULT: most genes change at 48 hours
# RESULT: the basemean is much lower for the genes that change at 96 hours

# Now let's add some bells and whistles to our bar plot!
```{r geom_bar()}
ggplot(filtered_res_df, aes(x = result_name, 
                 y = baseMean
                )) +
  geom_point() +
  stat_summary(fun = "mean",
               color = "red",
               shape = 1,
               size = 2)
```

# Summary of where we are at
Now we have gone through these layers:
1) Data -- data being plotted from object indexes etc.
2) MAPPING -- mapping data (aes)
3) Statistics -- statistical analyses on mapped data
4) Scales
5) Geom

# Let's revisit scales (layer 4) and facets (layer 6) briefly before moving on

4) scales -- how big small you want things
Scales is important for outputting specific factors as colors that are "scaled"

```{r more on scale layer}
# Let's see how we can use the scales layer to change axes more than just limits

ggplot(filtered_res_df, aes(x = pvalue, 
                            y =padj, color = padj < 0.01)) +
  geom_point()+
  scale_x_continuous() +
  scale_color_brewer(palette = "Set1") 

# scale_x_continious (same for y) is calling the scale layer specifically
# We are giving it where to put values on X-axis      
```

# changes axes limits
What if we want to set the limits of the axes? 
We can use 'xlim' & 'ylim' in scale layer.

```{R xlim and ylim}

ggplot(filtered_res_df, aes(x = pvalue, 
                            y =padj, color = padj < 0.0025)) +
  geom_point()+
  scale_x_continuous() +
  xlim(0, 0.01) +
  ylim(0, 0.01) +
  scale_color_brewer(palette =  "Set3") 
# coool so we zoomed in on the bottom left of the plot

```
