---
title: "03_plotting_ggplot_3"
author: "JR"
date: "2024-08-04"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(readr)
library(magrittr)
library(tibble)
library(dplyr)

```
# Load data
```{r loading our RNAseq analysis data TPM & COUNTS}
# load TPM RESULTS: 
load("/scratch/Shares/rinnclass/MASTER_CLASS/lessons/06_Differential_expression_analyses/results/TPM_results/TPM_results.Rdata")

# laod DESEQ2 results
load("/scratch/Shares/rinnclass/MASTER_CLASS/lessons/06_Differential_expression_analyses/results/counts_results_DESEQ/DESEQ_results.rdata")
```

# Summary of where we are at
Now we have gone through these layers:
1) Data -- data being plotted from object indexes etc.
2) MAPPING -- mapping data (aes)
3) Statistics -- statistical analyses on mapped data
4) Scales
5) Geom
# Now we are ready for the last three layers: FACETS, COORDINATE and THEME
6) Facet
7) Coordinate
8) Theme

# FACETS
This will allow us to make multiple plots in the same plot frame or facet
One limitation is that it can only make the same plots from different data

Be careful in the facet layer that you don't make too many plots. 
Essentially we make as many figures as there are entries.

# FACET_GRID & FACET_WRAP
```{r facet_grid(), facet_wrap()}

# Now let's pick a few genes of interest to plot.
genes <- c("Gm13698", "Gm13691", "Gm9045", "Pgk1-rs7", "Firre", "Neurod1")
gene_values <- filtered_res_df %>%
  filter(gene_name %in% genes)

# Now the plotting beings
ggplot(gene_values, 
       aes(x = log2FoldChange, y = padj )) +
  facet_grid(gene_name ~ .) + 
  geom_point()

# So facet grid makes each gene a row 
# Compare to facet wrap - looks better :)
ggplot(gene_values, 
       aes(x = log2FoldChange, y = padj )) +
  facet_wrap(gene_name ~ .) + 
  geom_point()
# facet wrap allows rows and columns in 

```
# Note that Firre was not in the list so wasn't plotted - but everything worked !
We can also facet grid another parameter -- let's say time_point

```{R two facets}

# let's facet by gene and time point 
ggplot(gene_values, 
       aes(x = log2FoldChange, y = padj )) +
  facet_wrap(gene_name ~ result_name) + 
  geom_point()

# Nice that worked and now we have a single data point in each plot :)
# This maybe a good example for facet grid:

genes <- c("Gm13698", "Gm13691", "Gm9045", "Pgk1-rs7", "Neurod1")
gene_values <- filtered_res_df %>%
  filter(gene_name %in% genes)

ggplot(gene_values, 
       aes(x = log2FoldChange, y = padj )) +
  facet_grid(gene_name ~ result_name) + 
  geom_point()

# NICE !

```

# DEEPER INTO FACET: we can add some parameters to order the plots
```{R deeper into facets}

# We can set the rows and cols in facet wrap
ggplot(gene_values, 
       aes(x = log2FoldChange, y = padj )) +
  facet_wrap(gene_name ~ ., nrow = 2, ncol = 3) + 
  geom_point()

?facet_wrap

# Note the axis are all the same for each plot -- let's use 'scales = free' parameter
ggplot(gene_values, 
       aes(x = log2FoldChange, y = padj )) +
  facet_wrap(gene_name ~ ., nrow = 2, ncol = 3, scales = "free") + 
  geom_point()
  
# Nice now every plot is scaled according to data in the plot !
```

# Coordinate layer
How to represent X and Y coordinates in different coordinate systems.
This is unlikely to really be used in biological data but an important layer!
We have been using Cartesian and it is the most common "coordinate" system.
But you can also plot data in "polar" and other coordinate systems.
We can just add it in the "Coordinates layer"

# To complete this layer let's just take a peak at "polar" coordinate space
```{r coordinate layer}

# let's make the coordinates polar instead of cartesian
ggplot(filtered_res_df, aes(x = result_name)) +
  geom_bar() +
  coord_polar()

# the values eminate from the center and are a "bar_plot" from the center.
# we can easily see as we did above that 24 and 48 hours had biggest change

# We can play around in the coordinate space too. 
# For example to tell what is  the "theta" or circular dimension. 

ggplot(filtered_res_df) +
  geom_bar(aes(x = result_name)) +
  coord_polar(theta = 'y')

# This helps us see that 12 and 96 had least amount of change
```

# Theme layer
This is a great way to package all the features you want, colors,
backgrounds, grid lines etc. We tend to use paper white as a defualt theme.

We made a defualt theme that we can source at the begining of the
document call _setup.R. Once loaded a "paper white" theme will be available.

Here is an example of a 'theme' is called in ggPlot
# There are lots of themes you can find them here:
https://ggplot2.tidyverse.org/reference/ggtheme.html

```{r themes titles}

ggplot(filtered_res_df, aes(x = log2FoldChange)) +
  geom_density(alpha = 0.2, color = "#424242", fill = "#424242") +
  theme_grey() +
  xlab(expression("log2fold change")) +
  ylab(expression("Density")) +
  ggtitle("log2fold change",
          subtitle = "genes regulated by dox and time")

# Note this example build everything from labels to data presentation!
# ylab and xlab labels X,Y azis
# ggtitle and subtitle etc 

# Let's take a look at _setup.R
# Nice, it basically calls a "function" in ggplot
# You can change this as you please by going through each line (hint: google)
```

# Let's make one more full plot with padj vs pvalue
```{r plotting values for linear regression on plot}
library(broom)

ggplot(filtered_res_df, aes(pvalue, padj)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  geom_text(
    x = mean(filtered_res_df$pvalue),
    y = max(filtered_res_df$padj),
    label = paste("y =", round(coef(
      lm(
        filtered_res_df$pvalue ~ filtered_res_df$padj,
        filtered_res_df$pavlue
      )
    )[2], 2), "* x +", round(coef(
      lm(
        filtered_res_df$pvalue ~ filtered_res_df$padj,
        filtered_res_df
      )
    )[1], 2)),
    parse = TRUE,
    color = "red",
    size = 6
  ) +
  ggtitle("Super cool linear regression?", subtitle = "pvalues are different than padj")

# Note that we did the lm in the "label" layer of geom_text()
# You can run funcitons pretty much anywhere in ggplot - custom figures !


```

# Custom themes
You can see that a theme is just a bunch of pre-bundled code!
https://ggplot2.tidyverse.org/reference/ggtheme.html

So when you arrive at your favorites you can save a lot of time 
by simplying sourcing a theme. Here is my favorite :) !

# Let's load and use a custom theme - paper white
Check out the file in "util/plot_theme.R"
It's an rscript that we source and then get all the parameters
```{r custom theme paperwhite}

source("../../util/plot_theme.R")

ggplot(filtered_res_df, aes(pvalue, padj)) +
  geom_point() +
  theme("paperwhite")+
  geom_smooth(method = "lm", se = FALSE) +
  geom_text(
    x = mean(filtered_res_df$pvalue),
    y = max(filtered_res_df$padj),
    label = paste("y =", round(coef(
      lm(
        filtered_res_df$pvalue ~ filtered_res_df$padj,
        filtered_res_df$pavlue
      )
    )[2], 2), "* x +", round(coef(
      lm(
        filtered_res_df$pvalue ~ filtered_res_df$padj,
        filtered_res_df
      )
    )[1], 2)),
    parse = TRUE, color = "red", size = 6
  ) 

```

# Very nice we can now use that them on any plot we have made previously let's check it out
```{r more theme paper white}

ggplot(gene_values, 
       aes(x = log2FoldChange, y = padj )) +
  theme("paperwhite", "col_pal10") +
  facet_grid(gene_name ~ .) + 
  geom_point()

# Same as before but now the lines are missing.
# We called a color pallete but it wasn't needed :)

```




