---
title: "05_heatmaps"
author: "JR"
date: "2024-08-13"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reshape)
library(tidyr)
library(tibble)
library(ggplot2)
library(dplyr)
library(pheatmap)
source("../../util/plot_theme.R")



```

# Goal create "heatmaps"to represent gene abundance across time and replicates.
Heatmaps are a very common and powerful way to represent RNAseq data. It shows the 
abundance of each gene, across all samples and shades them in colore
for more or less abundant expression (e.g., red high, blue low). Moreover, a heatmap can
be combined with a "clustering" algorithm to organize the patterns in the data. Finally, 
this also tells you how related samples are. We can do this in two steps:

# (1) cluster our samples to find the relationship across samples
# (2) cluster and overlay heatmap of gene abundance across time of dox exposure

Let's get started with "clustering our samples"
First let's load the data

```{r Loading data}

# load TPM RESULTS: 
load("/scratch/Shares/rinnclass/MASTER_CLASS/lessons/06_Differential_expression_analyses/results/TPM_results/TPM_results.Rdata")

# laod DESEQ2 results
load("/scratch/Shares/rinnclass/MASTER_CLASS/lessons/06_Differential_expression_analyses/results/counts_results_DESEQ/DESEQ_results.rdata")

```

# (1) Cluster samples
First step is calculating distance between all samples. Let's start with our
TPM data as this is typically  
# Note use of transmute (t) 
```{r distance calculation}

# Log-transform the TPM values
log_tpm_matrix <- log2(TPM_filtered + 1)

distance_matrix <- dist(t(log_tpm_matrix), method = "euclidean")

hc <- hclust(distance_matrix, method = "ward.D2")

plot(hc, labels = colnames(log_tpm_matrix), main = "Global TPM Filtered Dendrogram ", sub = "sample relationship by time and replicate after dox treatment")


# Cut tree
clusters <- cutree(hc, k = 5)
print(clusters)


#TODO auto ordered clustering


```


# (2) Overlapy a heatmap to measure sample similarity and gene changes across time
# PHEATMAP is a great program for making heat maps it will do 
```{r Create Heat map of all filtered TPM genes across all time points and replicates}

# THIS TAKES 3 MIN - but worth it 
pheatmap(log_tpm_matrix, 
         cluster_rows = TRUE,  # Cluster genes (rows)
         cluster_cols = TRUE,  # Cluster samples (columns)
         scale = "row",        # Scale the data across rows (genes) for better visualization
         show_rownames = FALSE, # Hide gene names for clarity
         show_colnames = TRUE,  # Show sample names
         main = "Heatmap of Log2-Transformed TPM Values for Sig genes")

#TODO include as bonus? 

custom_dist <- function(x) dist(x, method = "manhattan") # Replace "manhattan" with your preferred method
custom_hclust <- function(x) hclust(x, method = "ward.D2") # Replace "ward.D2" with your preferred clustering method

# Create a heatmap with custom distance and clustering functions
pheatmap(log_tpm_matrix, 
         cluster_rows = TRUE,  # Cluster genes (rows)
         cluster_cols = TRUE,  # Cluster samples (columns)
         scale = "row",        # Scale the data across rows (genes) for better visualization
         show_rownames = FALSE, # Hide gene names for clarity
         show_colnames = TRUE,  # Show sample names
         main = "Heatmap of Log2-Transformed TPM Values",
         distfun = custom_dist,    # Apply custom distance function
         hclustfun = custom_hclust # Apply custom clustering function
)


```


# Let's filter to sig genes
```{r}

sig_log_tpm_matrix <- log_tpm_matrix[rownames(log_tpm_matrix) %in% unique_sig_genes, ]

sig_distance_matrix <- dist(t(sig_log_tpm_matrix), method = "euclidean")

hc_sig_distance_matrix <- hclust(sig_distance_matrix, method = "ward.D2")

plot(hc_sig_distance_matrix, labels = colnames(hc_sig_distance_matrix), main = "Global TPM Sig Genes Dendrogram ", sub = "sample relationship by time and replicate after dox treatment for sig genes")


```

# Counts / rlog counts







# EXTRA

```{r}

# Cool we see if we choose 5 bins each time point falls into a bin with replicates !

ordered_colnames <- sort(colnames(log_tpm_matrix))
ordered_log_tpm_matrix <- log_tpm_matrix[, ordered_colnames]
hc_ordered <- hclust(dist(t(ordered_log_tpm_matrix)), method = "ward.D2")

# Step 3: Plot the dendrogram
plot(hc_ordered, labels = ordered_colnames, main = "Global TPM Filtered Dendrogram", 
     sub = "Sample relationship by time and replicate after dox treatment")

ordered_hclust <- hc$labels[hc$order]


ordered_hclust <- hc$labels[hc$order]

plot(ordered_hclust, labels = colnames(log_tpm_matrix), main = "Global TPM Filtered Dendrogram ", sub = "sample relationship by time and replicate after dox treatment")


```






