---
title: "06_raw_data_visualization"
author: "JR"
date: "2024-08-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal to visualize raw data for significant genes change due to dox across time.
We have now done a bunch of analyses using statistics and other means to home in on
genes that are regulated by dox and when. However, we haven't really gotten to see the raw data
yet. This would be the counts and where they pile up across the entire genome !

Now we are going to visualize the raw data for our favorite genes from our analyses thus far.

# Integrative Genomics Viewer (IGV) 
IGV is a great tool for visualizing read counts across the entire genome - and its FREE !
Let's get started by downloading IGV.

# (1) Download IGV
Download for your specific computer here:
https://igv.org/doc/desktop/#DownloadPage/

# - click the version for your comp and follow download instructions
# PICK JAVA INCLUDED VESION!
The main IGV website for more information is here:
https://igv.org/doc/desktop/##_top

# (2) Transfer BigWig files from your NF_CORE RNA_Seq pipeline output
In the output directory of your NF_CORE RNAseq pipeline there is a folder called
"bigWig" these are summaries of the number of RNAseq reads aligned at a genomic position.
These are Big so it may take a minute.

# (3) Drag bigWig files into IGV
# IMPORTANT: Make sure you select human genome "Mouse mm10" 
This is the genome version we used to align our reads in NF_CORE RNAseq pipeline

# (4) Drag in BigWig Files and Voila you can search the genome !!
Autoscale by slecting all samples and right click on left bar to select "group auto scale"

THere are a lot more tips and tricks on this accompanying YouTube video:
```{r}
#TODO
```






