---
title: "02_peaks_and_promoters"
author: "JR"
date: "2024-11-21"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../../util/00_our_first_function_lesson.R")
```

# GOAL : Find peaks that are present in all samples
We will further home our overlap skills here while asking the question:
How many peaks overlap in all samples? We previously did one overlap between
two files but we have 12 comparisons to do - How do we do that??

# Let's start by setting up file path to ATAC peak files. 
```{r setting up information to load peak files as GRange List}

# sourcing import peaks function 
source("../../util/00_our_first_function_lesson.R")

# establishing peak path to the dir with MACS2 output peak files from NF_CORE ATACseq pipeline

peak_path <- "<your_atac_pipeline_output_dir>/bwa/merged_library/macs2/broad_peak"
# peak_path <- "/scratch/Shares/rinnclass/MASTER_CLASS/lessons/08_ATACseq_pipeline/00_run_pipeline/00_pipeline_run/pipeline_run/bwa/merged_library/macs2/broad_peak"

# creating a file list also needed for import_peaks function to get sample name associated with file
fl <- list.files(peak_path, full.names = TRUE, pattern = ".broadPeak")
```

# Run import_peaks function to get a list of ATAC peak files in GRanges format

```{r run import_peaks}

my_peaks <- import_peaks(consensus_file_path = peak_path)

# Nice so with a list of peak files and path to the folder we can load as many peak files as we need !
num_peaks <- sapply(my_peaks, length) %>% as.data.frame
view(num_peaks)

# printing to compare later to filtered to cannonical chromosomes  
print(num_peaks)

```

# Now let's find a set of peaks that are common to all the samples !
We previously looked at the overlap between WT_0 and KO_O now let's find the peaks that are in all files.
Let's go through each step and then 

# strategy: use a for loop to iteratively find overlaps ! 
Let's do this and start to build a function to find overlaps automatically!
First, let's learn some reproducibility measures of grabbing names from files to ensure we know what they are.
```{r all time point consensus peaks}

# making a list of peak files (same as above - starting from scratch)
fl <- list.files(peak_path, 
                 full.names=TRUE, pattern = ".broadPeak")

# Extracting sample name from file name
sample_name <- sapply(fl, function(y){
 y <-  str_extract(y, "(KO|WT)_control_\\d+")
  
})

# Now we have the needed inputs to make a function !
# Let's make a fucntion to create peaks in all input files.
# Also to count how many peaks are in each file (number of overlaps)
```

# Create for loop and function at the same time !!
Now we are going to create a function to find overlaps between
all files in a list of GRanges (in our case right now my_peaks).
Let's create a handy function to overlap peaks!

```{r find_common_peaks function }

# starting function named "find_common_peaks" parameter 
find_common_peaks <- function(gr_list) {
  # Validate input
  if (!is.list(gr_list) || !all(sapply(gr_list, inherits, "GRanges"))) {
    stop("Input must be a list of GRanges objects.")
  }
  
  # Start with the first GRanges object
  common_peaks <- gr_list[[1]]
  
  # Iteratively find overlaps across all GRanges objects
  for (i in 2:length(gr_list)) {
    current_gr <- gr_list[[i]]
    
    # Find overlaps
    overlaps <- findOverlaps(common_peaks, current_gr)
    
    # Subset to overlapping regions
    common_peaks <- subsetByOverlaps(common_peaks, current_gr)
  }
  
  # Assign custom names to the common peaks
  mcols(common_peaks)$name <- paste0("common_peak_", seq_along(common_peaks))
  
  return(common_peaks)
}

# Let's give it a test run !

common_peaks <- find_common_peaks(my_peaks)
common_peaks@elementMetadata@listData[["name"]]
```

# IMEDIATE TASK: Take a moment and make this function in your .Rscrip in util

(a) Add this function above to your .Rscript in util directory. 
(b) Add description to the .Rscript
(c) clean environment - source .Rscript then run with just this code:

```{r (c) code to test function}

# peak file "consensus_path" parameter of import_peaks function
peak_path <- "<your_atac_pipeline_output_dir>/bwa/merged_library/macs2/broad_peak"
# peak_path <- "/scratch/Shares/rinnclass/MASTER_CLASS/lessons/08_ATACseq_pipeline/00_run_pipeline/00_pipeline_run/pipeline_run/bwa/merged_library/macs2/broad_peak"

# source function
source("../../util/00_our_first_function_lesson.R")

# make fl
fl <- list.files(peak_path, full.names = TRUE, pattern = ".broadPeak")

# import peaks !
my_peaks <- import_peaks(consensus_file_path = peak_path)

# test common_peaks function 
common_peaks <- find_common_peaks(my_peaks)

```
NICE !! In the future that is all you will have to run (5 lines of code !!)
HINT: This is a good example of what your final .RMD could look like above - nice and clean!

# Now: let's prepare and view peak calls in IGV to do some analysis in the following steps:

# (1) We need to make our common peaks object into a .BED file to view in IGV.
Let's make a handy script to make peak objects in environment into viewable files !

```{r exporting common_peaks GRanges to .bed file for viewing in IGV}

# setting export file path
output_bed <- "results/common_peaks.bed"

# Export the GRanges object to a BED file
rtracklayer::export(common_peaks, con = output_bed, format = "BED")

# That's it now we can visualize the common_peaks you just created!
```
# (2) Now to explore and visualize our peak calls with raw data
You will need to use cyberduck/filezilla to transfer to your local computer:
(a) your bigWig files for each ATACseq file run (this may take 10 min or so)
  - hint(/bwa/merged_library)
(b) common_peaks.bed file just created.

# (3) Load in the common peaks and individual ATAC bigwig files to IGV and investigate:
(a) Do the peaks look like they are common in all samples?
(b) search around for good peaks and note some examples (the coordinates on IGV - put in .RMD)
(c) What is weird about Firre? 
(d) how many peaks are in your common_peaks file versus the one produced by NF_CORE pipeline?
  - Hint <output_dir> /bwa/merged_library/macs2/broadPeak/consensus/consensus_peaks.mLb.clN.boolean.intersect.plot.pdf

# (4) Now load in the consensus peak file that has any peak in any sample and compare in IGV
(a) Find a region where there is a common peak next to a peak that is not common but called by pipeline
(b) Look at some common peaks and see how often a consensus_peak is present - would expect almost all right?
(c) what is the difference you notice between the boundary length of consensus -vs- common? Are the peaks called bigger or smaller?

# Congratulations you just completed one of the most important tasks in computational biology:
Running scripts to find overlaps across samples and visualizing what that looks like in raw data!
This is useful in almost all applications - always check back to raw data - well done !!


