---
title: "03_Peak_genome_features"
author: "JR"
date: "2024-11-24"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../../util/00_our_first_function_lesson.R")
```

# Goal: To determine where in the genome our peaks live!
We will start to explore where our ATACseq peaks reside in the human genome
and what features they correspond to (e.g., mRNAs, lincRNA and many more to explore)

# Note difference between the two peak files we have been using:
# common peaks: represent peaks present in all samples (time-points).
# consensus peaks: peaks present in any sample(time-point) - from NF_CORE MACS2 
So for now we are investigating where are the regions of chromatin
that remain accessible despite dox exposure across any time point.

This is a really cool question of where do these robust and reliable peaks reside?

# Let's find out where in the genome our peaks are through these steps:

# Step (1) : create "common_peaks" a list of peak files in the format of GRanges
We are basically starting where we left off last lesson!

```{r create my_peaks GRange List using import_peaks custom function}

# peak file "consensus_path" parameter of import_peaks function
peak_path <- "<your_atac_pipeline_output_dir>/bwa/merged_library/macs2/broad_peak"
#peak_path <- "/scratch/Shares/rinnclass/MASTER_CLASS/lessons/08_ATACseq_pipeline/00_run_pipeline/00_pipeline_run/pipeline_run/bwa/merged_library/macs2/broad_peak"

# make file list (fl)
fl <- list.files(peak_path, full.names = TRUE, pattern = ".broadPeak")

# import peaks !
my_peaks <- import_peaks(consensus_file_path = peak_path)

# test common_peaks function 
common_peaks <- find_common_peaks(my_peaks)

```
Sweet! We are all set with peak files/objects we need for our analyses !

# now for genomic annotation GRanges

Here we will import the gencode genome annotation (for mm10) as a GRanges.
Then we will subset it to "genes", "mRNA", "lncRNA" 
This will allow us to determine how many common_peaks are 
near these genome features. Then the same concept applies to all genome features!
# In short: we are making GRanges of genome features to use in overlaps with ATAC peaks

We have done this before but will start from scratch to refresh:
# Step (2) Creating mouse gene, lincrna, mRNA annotation GRange objects

```{r create mouse genome annotation for genes, lincRNA, mRNA in GRanges format}

# Loading gencode genome annotation as GRanges (takes 5 min - will save at end)
gencode_gr_mouse <- rtracklayer::import("/scratch/Shares/rinnclass/MASTER_CLASS/GENOMES/M25/gencode.vM25.annotation.gtf")

# making subset to just gene objects:
gencode_genes_mouse <- gencode_gr_mouse[gencode_gr_mouse$type == "gene"] 

# mrna_genes
mrna_genes <- gencode_genes_mouse[gencode_genes_mouse$gene_type %in% "protein_coding"]

# lincrna_genes
lincrna_genes <- gencode_genes_mouse[gencode_genes_mouse$gene_type %in% "lincRNA"]

```
# Result: 55,401 genes, 21,895 mRNA, 5,629 lincRNAs in mouse genocde annotation 
# Result this means that half of the gene annotations are in other catagories such as "processed_transcripts"

Cool, so we have learned more about the number of genes in the mouse genome.

# Step (3) : Promoters object from TSS to upstream and downstream
This is very subjective to what "window" you create a promter around a TSS. 
So we will start with a typical window to learn, while analyzing our ATACseq data!
We did this previously in 08_ATACseq_Pipeline/05_ & 06_ 

A good reference window is 2Kb upstream and 2Kb Downstream for transcriptional regulatory events.
Let's make a promoter window around each of the gene types we just loaded !

```{r Creating promoter GRanges for genes, lncrna, mrNA}

# mrna_promoters
mrna_promoters <- promoters(mrna_genes, upstream = 2000, downstream = 2000)

# gene_promoters
gene_promoters <- promoters(gencode_genes_mouse, upstream = 2000, downstream = 2000)

# lncRNA gene_promoters
lincrna_gene_promoters <- promoters(lincrna_genes, upstream = 2000, downstream = 2000)

# Nice now we have a lot of Genome Features in GRanges format - ready to start overlapping.
```

# Now we are set to start overlapping.
Let's ask several questions to explore the data and learn some skills !

# Question 1: How many common peaks overlap mRNA vs lincRNA promoters
```{r ATAC Peak Overlap lincRNA and mRNA promoters}

lincrna_atac_ov <- findOverlaps(lincrna_gene_promoters, common_peaks)

length(lincrna_gene_promoters)
# 5,629 entries

length(lincrna_atac_ov)
# 1,073 overlaps about 20% of lincrna promters have atac peaks in all samples.
(1073)/(5629)

# let's check our overlap numbers:
length(lincrna_atac_ov@to)
length(lincrna_atac_ov@from)
# Cool all looks good !


# Now let's check mRNA promoter overlaps# let's check our overlap numbers:
mrna_atac_ov <- findOverlaps(mrna_promoters, common_peaks)
length(mrna_atac_ov)
length(mrna_atac_ov@to)
length(mrna_atac_ov@from)
# 13,700 overlaps ! That's a lot !

# checking percentage of promoter overlaps
(13700)/(21859)
# also
(length(mrna_atac_ov))/ ((length(mrna_genes)))
# 63% of mRNA promoters have overlaps with an ATAC peak in all samples!
# NOTE not exactly the same which is strange .... 61.2% vs 62.6% 
# That is way more than lincRNA. This raises quesiton of which ones were expresses?
# We will get to that later.

# percentage of common peaks with promoter overlaps 
(21859 + 1073) / (37152)
# 62% ! That's a lot consider how small this space is overall in the genome
```
# Result: 63% of mRNA promoters have ATAC peak promoter overlap 
# Result: 20% of lincRNA promoters have ATAC peak promoter overlap
# Result: Total of ~22,500 out of 37,152 common peaks 
# Result: Stable ATAC peaks in all samples are biased to promtoers 

# What is the percentage of overlaps with all genes?
```{r all gene overlap with common peaks}

gene_promoter_ov <- findOverlaps(gene_promoters, common_peaks)
length(gene_promoter_ov)
length(gene_promoter_ov@to)
length(gene_promoter_ov@from)

# That is 19,126 overlaps of 55,401 gene annotations
(19126)/(55401)
# 35% ! Lower than mRNAs alone !
```
# Result: 35% of gene promoters have ATAC peak overlaps in all samples
# Result: This is 2 times lower than what we observed for mRNA
# Result: mRNA promoters are highly enriched in commone mouse ESC ATAC sites


# Now let's compare the overlaps from our "find_common_peaks()" custom function
```{r finding overlaps with "find_common_peaks" function}

# need list of GRanges as input so merging these two compare each other
gr_list <- list(mrna_promoter = mrna_promoters, common = common_peaks)

# finding common peaks
mrna_atac_ov_common <- find_common_peaks(gr_list)
length(mrna_atac_ov_common)
# Ok we see 13,316 overlaps compared to 13,700 using findOverlaps

# Let's see if find_overlaps approach and common overlap.
# strategy put them all in a list and then use common overlap 

gr_list2 <- list(sample1 = mrna_promoters, sample2 = common_peaks, sample3 = mrna_atac_ov_common)

# running find common
mrna_atac_ov_common2 <- find_common_peaks(gr_list2)
length(mrna_atac_ov_common2)
# Same 13,316 - this means all common were present in findOverlaps
```
Nice we have analyzed and found best strategy for finding common peaks.

# Now let's nicely repeat the process for gene, lincRNA and mRNA !
Next we want to visualize and make sure peaks are correctly called on mRNA and lncRNA promoters
First we need to do the common_peaks function to get all the objects in right format.
Then make a bed file below : almost there let's go !

# (a) GENE overlaps: common_peak ATAC peaks and gene promoter overlaps
# IMPORTANT : PEAKS need to come first !!
```{r using common_peaks function to create objects to be made into bed files}

# make a list of gene_promoters and common_peaks
gr_list_genes <- list(sample1 = common_peaks, sample2 = gene_promoters)
# find common overlaps
all_gene_common_ov <- find_common_peaks(gr_list_genes)
length(all_gene_common_ov)
# 14,761 smaller than 19,126 because of ordering!

# create a bed file for peaks:
rtracklayer::export(all_gene_common_ov, con = "results/gene_common_ov.bed", format = "BED")
```

# (b) lincRNA promoter ATAC overlap bed
```{r lincrna promoter common overlaps and bed file}

# make a list of lincrna_promoters and common_peaks
gr_list_genes <- list(sample1 = common_peaks, sample2 = lincrna_gene_promoters)
# find common overlaps
lincrna_common_ov <- find_common_peaks(gr_list_genes)
length(lincrna_common_ov)
# 1035 about the same 

# create a bed file for peaks:
rtracklayer::export(lincrna_common_ov, con = "results/lincrna_common_ov.bed", format = "BED")
```

# (c) mrna promoter ATAC overlap
```{r mrna promoter common overlaps and bed file}

# make a list of lincrna_promoters and common_peaks
gr_list_genes <- list(sample1 = common_peaks, sample2 = mrna_promoters)
# find common overlaps
mrna_common_ov <- find_common_peaks(gr_list_genes)
length(mrna_common_ov)
# 12,000 smaller but not too far off

# create a bed file for peaks:
rtracklayer::export(mrna_common_ov, con = "results/mrna_common_ov.bed", format = "BED")

```

# (d) let's save our data / session !
We will need some key objects for future analyses so let's save!
```{r saving env objects}

save(lincrna_genes, lincrna_gene_promoters, mrna_genes, mrna_promoters, gene_promoters, gencode_genes_mouse, common_peaks, mrna_common_ov, my_peaks, lincrna_common_ov, all_gene_common_ov, file = "results/key_mouse_atac_data.Rdata" )
```


# (e) Now let's take a look at our peak overlaps in IGV (common practice !)
First we need to make some handy bed files of the overlaps we did.

# Let's ask the following questions about our peak files (feel free to take notes here!):
First we need to load all 4 .bed files we have created and all the bigWig files into IGV.
Also please load the consensus peaks file we used last time (any peak in any sample)

# Questions to explore:
(a) Find a common peak that is also called as mRNA
  - how do these peaks look? Where do they appear to be relative to nearby consensus peaks?

(b) Find a common peak that is also called a lncRNA
   - how do these peaks look? Where do they appear to be relative to nearby consensus peaks?
   
(c) Do any peaks get called as both lincrna and mRNA?
  - do this in code as well :) !

(d) How could any lncRNA and mRNA promoter regions both be called as overlaps?
  # Hint BC1
  

