---
title: "01_run_DESEQ2"
author: "JR"
date: "2024-04-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(IRanges)
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(purrr)
library(magrittr)
library(pheatmap)
library(textshape)
library(Rcpp)
library(DESeq2)
```

# Goal to run DESEQ2 and get familiar with results
# Please note that we need to organize our new directory with "analysis, figures and results directories"
Using DESEQ2 we will determine which genes change in RNA expression levels
at different times of dox exposure. Let's officially start doing some analysis!

We are going to start slowly and stepwise to get a feel for DESEQ2.
It is highly recommended that you read these documents for deeper dives!

https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html


# Let's start by loading in our counts_matrix and sample info
# IMPORTANT: careful of order when loading Rdata :)
```{r loading in counts and sample info}

# /scratch/Shares/rinnclass/MASTER_CLASS/lessons/05_RNAseq_in_R_studio/results
load("../05_RNAseq_in_R_studio/results/deseq_samples.RData", verbose = T)

# NOte the write over for counts matrix and ordering of loading these two objects
load("../05_RNAseq_in_R_studio/results/count_files.RData", verbose = T)

```

# (1) first run DESeq2 by creating a dds (DESEQ Data Structure) 
We use the function "DESeqDataSetFromMatrix"
With parameters of countData, colData, design

dds <- DESeqDataSetFromMatrix(countData = counts_filtered,
                              # this is our counts data we read in
                              colData = desseq_samples,
                              # telling DeSeq what env variable to use for sample sheet
                              design = ~ condition)
                              # perhaps most important is "condition" is a factor in deseq_samples
                              # we will be using time_point
                              
# Ok let's make a dds !

```{r run DEseqDataSetFromMatrix }

dds <- DESeqDataSetFromMatrix(countData = counts_filtered,
                              colData = deseq_samples,
                              design = ~ time_point)
```

# Now run DESeq function that does all the magic !
```{r run DEseqDataSetFromMatrix }

dds <- DESeq(dds)

```

# Not only was that fast but numerous analyses were just performed !
Let's take a look at the result outputs
Important: to see which analyses DESEQ2 performed 

```{r results}

# resultsNames function retreives names of results
resultsNames(dds)

# Nice we see this output:
# "Intercept" "time_point_12_vs_0" "time_point_24_vs_0" "time_point_48_vs_0" "time_point_96_vs_0"
# DESEQ found genes differnetial expresion values for each time point relative to 0 time point.


```
# RESULT: we have compared gene expression values relative to 0 timepoint.
It is good to start making a notation of results you wan to keep track of.
In this case it is key to record what analysis was done.

# Next:Running DeSeq to create rlog normalzied counts!
Although a lot of normalization has already occurred from
Salmon (where we got the counts) we now need to determine the 
distribution of expression values and normalize again to prevent
low abundance genes inflating significant differential expression. 


#  Normalize counts (rlog function)
This basically is rank counts normalized to std error in replicates.

```{r rlog }
rlog_counts <- rlog(dds, blind = TRUE)
```

# now we retrieve the values using the "assay" function that converts to rlog_counts
```{r}
rlog_counts_matrix <- assay(rlog_counts)

```

# Finally, let's save this rlog matrix to use in our future analyses and plotting.
We will save this as R Data Structure .rds -- this will keep the object stored properly
to be loaded into the environment to use for downstream analyses.

```{r}

write_rds(rlog_counts_matrix, "results/rlog_counts_mouse_dox_long.rds")

```

# Nice you now have run DESEQ2! 
There is a lot more to explore
In the meantime enjoy all the analyses DESEQ2 just did for you !