---
title: "02_DESEQ2_start_analysis"
author: "JR"
date: "2024-04-21"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DESeq2)
library(readr)
library(magrittr)
library(tibble)
library(dplyr)
library(ggplot2)

#BiocManager::install("apeglm")

```

```{r loading data}
load("../05_RNAseq_in_R_studio/results/count_files.RData", verbose = T)
load("../05_RNAseq_in_R_studio/results/deseq_samples.RData", verbose = T)
```

# Long mouse dox time course DDS
```{r run DESEQ2}
# create DDS
mouse_long_dox_dds <- DESeqDataSetFromMatrix(countData = counts_filtered,
                                                 colData = deseq_samples,
                                                 design = ~ time_point)


# Run DESEQ2 on DDS
mouse_long_dox_dds <- DESeq(mouse_long_dox_dds)
```

# Let's look at the results
```{r result names from DESEQ2}

resultsNames(mouse_long_dox_dds)

# Cool, we see a bunch of results of time_point_X_vs_0 
# Those are our results !
# We just don't want the intercept file
```

# Now let;s make result files for 0 hours versus 12 hours 
The goal is to learn some key concepts from DESEQ2 results() function
Then we will move on to comparing across all time points.
But first let's get down to some fundamental analyses.

```{r extracting results from DDS}
# fist let's see how results work
# contrast = c("time_point", "12", "0") specifies that you want to compare 
# the "12" level of the "time_point" factor against the "0" level. 
# Antoher reason why factoring our DESEQsamples file is importnat

res_0 <- results(mouse_long_dox_dds, contrast = c("time_point", "12", "0"))
view(res_0)

# IMPORTANT: "time_point" here is the factored column and would be different
# if you called this column differently in your sample file (deseq_samples)
```

# Here's a breakdown of the arguments above:
"time_point" is the name of the factor (column) in your count data that you want to compare.
"12" is the level of the factor that you want to compare against.
"0" is the reference level of the factor (i.e., the baseline).


# res_0 is an S4 object and abit confusing let's turn into a data frame:
```{r making a dataframe of our results}
res_0_df <- data.frame(time = 12, res_0)
view(res_0_df)

# That is a bit better - as a dataframe
# But out gene_id is still in the "meta" 
# lets move it into a column with rownames_to_column

res_0_df <- rownames_to_column(res_0_df, "gene_id")
view(res_0_df)
```

# We can notice that there are NAs in the padj column let's see how many with is.na() function
```{r checking NAs}

# counting NAs
sum(is.na(res_0_df$padj))

# How many genes in DF
nrow(res_0_df)

# what percentage
sum(is.na(res_0_df$padj)) / nrow(res_0_df)

# Yikes half our genes have NA for padj - let's deal with that later 
# In the meantime is.na is really important way to check your data 
# for character values that maybe intermingled with numeric - a problem
# let's ignor for now :)
```


# Cool, now let's merge in the g2s to get the gene names
We made a g2s and this is where it comes in handy !
```{r merge in gene_name}
# we need to merge g2s and res_0_df with a column that exists in both DFs = gene_id

res_0_df <- merge(res_0_df, g2s, by = c("gene_id" = "gene_id"))
view(res_0_df)
# Or we could use left join or join 
res_0_df <- left_join(res_0_df, g2s)
view(res_0_df)
# Let's orgnaize this so the gene name is next to gene_id
res_0_df <- res_0_df[, c(1, 9, 2:8)]
view(res_0_df)
# cool now we can actually start to do some analysis for the 
#  0 vs 12 hour time point.

```

# Now let's practice filtering this data frame on padj adn lfcSE
This is a good opportunity to learn some tidyverse with the pipe
%>% function
```{r filtering our results to significant Padj < 0.05}

# filter() function for padj
# note that %>% pushes data to next function (filter in this case)
filtered_res_0_df <- res_0_df %>%
  filter(padj < 0.05)

nrow(filtered_res_0_df)
# cool now we see 720 genes with padj < 0.05

# let's filter on both padj and log2FoldChange
# these arguments are simply added by a , and could have even more !
filtered_res_0_df_2 <- res_0_df %>%
  filter(padj < 0.05, abs(log2FoldChange) > 1)

nrow(filtered_res_0_df_2)

# Interesting this is now only 42 genes !! 
# That is a special number :)

```

# let's take a look at what "type of genes" are changing at 12 hours.
# Let's use res_0_df_2

We can extract the gene names and then do gene ontology
```{r extracting sig genes at P <0.05 and LFC > abs(1)}

# printing out the gene_symbol column 
cool_genes <- cat(paste(filtered_res_0_df_2$gene_symbol, collapse = "\n"))

cool_genes_12 <- as.data.frame(filtered_res_0_df_2$gene_symbol, collapse = "\n")

# copy and paste into EnrichR 
#https://maayanlab.cloud/Enrichr/


```
# Interesting, seems like our genes are involved in mitochonrial functions
Nice, we just started analyzing significant genes that respond to dox at 12 hours !!

We will dive much deeper into plotting as we go, but for now a standard first plot
is a volcano plot. Let's make one !

# volcano plot of genes significant at 12 hours
```{r for fun making a volcano plot}

library(ggplot2)

ggplot(filtered_res_0_df_2, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point() +
  theme_classic() +
  labs(x = "Log2 Fold Change", y = "-Log10 Adjusted P-value") +
  geom_hline(yintercept = -log10(0.05), color = "red") +
  geom_vline(xintercept = c(-1, 1), color = "blue")

```
# congratulations you have started your fisrt RNAseq analyses !
# Let's dig in more in the next lesson

###################
Bonus Exercise 
###################

# let's do the same thing for 0 vs 96 hours our latest time point.
```{r}

res_96 <- results(mouse_long_dox_dds, contrast = c("time_point", "96", "0"))
res_96_df <- data.frame(time = 96, res_96)
res_96_df <- rownames_to_column(res_96_df, "gene_id")
res_96_df <- merge(res_96_df, g2s, by = c("gene_id" = "gene_id"))

filtered_res_96_df <- res_96_df %>%
  filter(padj < 0.05)

filtered_res_96_df_2 <- res_96_df %>%
  filter(padj < 0.05, abs(log2FoldChange) > 1)

cool_genes_96 <- as.data.frame(filtered_res_96_df_2$gene_symbol, collapse = "\n")

```

# Plotting
```{r}

library(ggplot2)

ggplot(filtered_res_96_df_2, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point() +
  theme_classic() +
  labs(x = "Log2 Fold Change", y = "-Log10 Adjusted P-value") +
  geom_hline(yintercept = -log10(0.05), color = "red") +
  geom_vline(xintercept = c(-1, 1), color = "blue")


```


# determining overlap
```{r}
overlap_genes <- intersect(cool_genes_96, cool_genes_12)

overlap_genes <- cool_genes_12 %in% cool_genes_96
```
