---
title: "02_DESEQ2_start_analysis"
author: "JR"
date: "2024-04-21"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DESeq2)
library(readr)
library(magrittr)
library(tibble)
library(dplyr)
library(ggplot2)

BiocManager::install("apeglm")

```

```{r}
load("../05_RNAseq_in_R_studio/results/count_files.RData", verbose = T)
load("../05_RNAseq_in_R_studio/results/deseq_samples.RData", verbose = T)
```

# Long mouse dox time course DDS
```{r}
# create DDS
mouse_long_dox_dds <- DESeqDataSetFromMatrix(countData = counts_filtered,
                                                 colData = deseq_samples,
                                                 design = ~ time_point)


# Run DESEQ2 on DDS
mouse_long_dox_dds <- DESeq(mouse_long_dox_dds)
```

# Let's look at the results
```{r}

resultsNames(mouse_long_dox_dds)

# Cool, we see a bunch of results of time_point_X_vs_0 
# Those are our results !
# We just don't want the intercept file
```

# Now let;s make result files for each time point comparision using results() function

Here's a breakdown of the arguments:
"time_point" is the name of the factor (column) in your count data that you want to compare.
"12" is the level of the factor that you want to compare against.
"0" is the reference level of the factor (i.e., the baseline).
```{r}
# fist let's see how results work
# contrast = c("time_point", "12", "0") specifies that you want to compare 
# the "12" level of the "time_point" factor against the "0" level.
res_0 <- results(mouse_long_dox_dds, contrast = c("time_point", "12", "0"))

# it's an S4 object and abit confusing let's turn into a data frame:

res_0_df <- data.frame(time = 12, res_0)

# That is a bit better - as a dataframe
# But out gene_id is still in the "meta" 
# lets move it into a column with rownames_to_column

res_0_df <- rownames_to_column(res_0_df, "gene_id")

# Cool, now let's merge in the g2s to get the gene names
# we need to merge with a column that exists in both DFs = gene_id

res_0_df <- merge(res_0_df, g2s, by = c("gene_id" = "gene_id"))

# cool now we can actually start to do some analysis for the 
#  0 vs 12 hour time point.
# We see about 700 genes that are significantly differential between 0 and 12 hours !
```

# Now let's practice filtering this data frame on padj adn lfcSE
```{r}

# filter() function for padj
filtered_res_0_df <- res_0_df %>%
  filter(padj < 0.05)

# cool now we see 720 genes with padj < 0.05

# let's filter on both padj adn log2FoldChange
filtered_res_0_df_2 <- res_0_df %>%
  filter(padj < 0.05, abs(log2FoldChange) > 1)

# Interesting this is now only 42 genes !! 
# That is a special number :)

```

# let's take a look at what "type of genes" are changing at 12 hours.
# Let's use red_0_df_2

We can extract the gene names and then do gene ontology
```{r}

# printing out the gene_symbol column 
cool_genes <- cat(paste(filtered_res_0_df_2$gene_symbol, collapse = "\n"))

# copy and paste into EnrichR 
#https://maayanlab.cloud/Enrichr/


```
# Interesting, seems like our genes are involved in mitochonrial functions
Nice, we just started analyzing significant genes that respond to dox at 12 hours.

We will dive much deeper into plotting as we go, but for now a standard first plot
is a volcano plot. Let's make one !

# volcano plot of genes significant at 12 hours
```{r}

library(ggplot2)

ggplot(filtered_res_0_df_2, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point() +
  theme_classic() +
  labs(x = "Log2 Fold Change", y = "-Log10 Adjusted P-value") +
  geom_hline(yintercept = -log10(0.05), color = "red") +
  geom_vline(xintercept = c(-1, 1), color = "blue")

```

# Note that there is only one value for each gene but we had three replicates?
That is becuase when we use results() function we get the mean across replicates 
if we want all the values we need to use counts
```{r}
# First let's see what are column names are for each replicate:
colnames(counts(mouse_long_dox_dds, normalized = FALSE))
# They all start with WT_0

# cool now we can grep all the 0 timepoint values using "^WT_0_"
counts_0 <- counts(mouse_long_dox_dds, normalized = TRUE)[, grep("^WT_0_", colnames(counts(mouse_long_dox_dds, normalized = TRUE)))]

# Let's grab the counts for 12 hour time point as well
counts_12 <- counts(mouse_long_dox_dds, normalized = TRUE)[, grep("^WT_12_", colnames(counts(mouse_long_dox_dds, normalized = TRUE)))]

# Nice now we have an object of each replicate for 0 hour counts and 
res_counts <- results(mouse_long_dox_dds, contrast = c("time_point", "12", "0"))
significant_genes_counts <- res_counts[res_counts$padj < 0.05, ]

# we see NAs let's get rid of them
res_counts <- res_counts[!(is.na)(res_counts$padj), ]

significant_genes_counts <- res_counts[res_counts$padj < 0.05, ]



plotMA(res, main = "MA Plot: 0 vs 12 hours", ylim = c(-5, 5))
points(significant_genes$baseMean, significant_genes$log2FoldChange, col = "red")

# Cool this gave us fold change versus expression level or MA plot
# This is a standard first plot and as usual the more expressed
# the less the fold change
```

# Now let's plot the standard deviation across replicates between 0 and 12 hours
```{r}
# we are going to subset using %in% the signifcant genes from the counts object with all three replicates.
counts_0_sig <- counts_0[rownames(significant_genes_counts) %in% rownames(counts_0), ]

# Same for 12 hours
counts_12_sig <- counts_12[rownames(significant_genes_counts) %in% rownames(counts_12), ]

```

# NOw we are going to calcualte the mean and standard deviation across replicates
```{r}
# mean
mean_0 <- apply(counts_0_sig, 1, mean)

# standard error
se_0 <- apply(counts_0_sig, 1, function(x) sd(x)/sqrt(length(x)))

# same for 12 hour counts
mean_12 <- apply(counts_12_sig, 1, mean)

# standar error 12 hours
se_12 <- apply(counts_12_sig, 1, function(x) sd(x)/sqrt(length(x)))
```

# Now plotting 
again we will go into this more just to see our results for now

```{r}
# Plot 0 vs 12 hours with error bars
library(ggplot2)

df <- data.frame(
  gene = rownames(significant_genes_counts),
  time_point = c(rep("0", length(mean_0)), rep("12", length(mean_12))),
  mean = c(mean_0, mean_12),
  se = c(se_0, se_12)
)

ggplot(df, aes(x = time_point, y = mean)) +
  geom_point(aes(color = gene), position = position_dodge(0.1)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, color = gene), position = position_dodge(0.1)) +
  theme_classic() +
  labs(x = "Time Point", y = "Normalized Counts")


```







