---
title: "04_exploring_results"
author: "JR"
date: "2024-07-15"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(IRanges)
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(purrr)
library(magrittr)
library(pheatmap)
library(textshape)
library(Rcpp)
library(DESeq2)
```

# GOAL : Let's explore our results data frame using the time point analysis from DESEQ2
We now have results after all that hardwork - let's see what we see and start to explore !

# Loading our results from DESEQ for all time points
```{r loading our DESEQ2 results res_df}
load("/scratch/Shares/rinnclass/MASTER_CLASS/lessons/06_Differential_expression_analyses/results/time_point_res_df.RData")
```


# First let's filter to find genes that significantly changed across time points
For this we will use the filter function in baseR


```{r filter time point to padj < 0.05}

filtered_res_df <- res_df %>%
  filter(padj < 0.05)

# Wow we can see a major reduction in our dataframe

```

# Now let's see what happens when we filter for fold change across time as well 
```{r filtering Padj < 0.05 and LFC abs(1)}

filtered_res_df_2 <- filtered_res_df %>%
  filter(padj < 0.05, abs(log2FoldChange) > 1)

```

# How many unique genes does this represent?
```{r how many unique genes are in the significant results df?}

sig_genes <- unique(filtered_res_df_2$gene_id)
# cool so we have 669 genes that are significantly changing in at least one time point
```
# Result (again keeping track of good results)
# Result: There are 669 unique genes that change across time (padj < 0.05 log2 > 1)

# Which gene has the lowest padj?
```{r exploring lowest padj}

lowest_pval <- min(filtered_res_df_2$padj)
# or

lowest_pval <- filtered_res_df_2$padj %>%
  min()

# Let's grab the whole row instead of just the value (note need to use which.min)
lowest_pval <- filtered_res_df_2[which.min(filtered_res_df_2$padj),]

# Cool what is this gen what does it do?
# go google for the gene name !
```

# Which gene has the largest fold change across any time point?
```{r searching for maximum fold change in results}

max_value <- filtered_res_df_2$log2FoldChange %>% 
  max()

# Let's get the whole row to get gene name etc.
max_row <- filtered_res_df_2[which.max(filtered_res_df_2$log2FoldChange),]

# Nice what is this gene? When did it go up this much 
```

# Lets do gene pathway analysis on the significant genes affected by dox across time
```{r getting gene list as data frame and then doing ontology analysis}
# printing out the gene_symbol column 
dox_genes <- cat(paste(filtered_res_df_2$gene_name, collapse = "\n"))

# as data frame:
dox_genes <- as.data.frame(filtered_res_df_2$gene_name, collapse = "\n")
# Let's make sure to only get the unique names
dox_genes <- unique(dox_genes)

# Renaming the column (handy code that is used a lot :)
dox_genes <- dox_genes %>% 
  rename(gene_name = 'filtered_res_df_2$gene_name')

colnames(dox_genes)

# Now let's write this out and do gene enrichment analysis
write.table(dox_genes["gene_name"], row.names = FALSE, col.names = FALSE, "results/gene_names.csv")

# copy and paste into EnrichR 
#https://maayanlab.cloud/Enrichr/

```

# let's take a look at base mean distribution of our sig genes
# Let's learn the hist function
```{r exploring the distribution of basemean and padj values}

hist(filtered_res_df_2$baseMean)

# Let's change the X-axis
hist(filtered_res_df_2$baseMean, xlim = c(0,1000), breaks = 1000)

# Let's do the same for Pvalue
hist(filtered_res_df_2$padj)

# Fold change
hist(filtered_res_df_2$log2FoldChange, xlim = c(-5, 5), breaks = 1000)

# How does this compare to Lfsce
hist(filtered_res_df_2$lfcSE)


```

# Now let's make some comparisons between our significant genes and all other genes

```{r finding median basemean values}

# Let's see the median basemean expression of all genes
basemean_all_genes <- median(res_df$baseMean)

# Let's do the same for our sign genes 
basemena_sig_genes <- median(filtered_res_df_2$baseMean)

# Ok looks a lot higher basemean for our sign genes.
# Let's do a t-test to see if this is signficant :

?t.test
t.test(res_df$baseMean, filtered_res_df_2$baseMean)

# Cool it's a significant result :) !

```
# Result: The dox response genes have a higher basemean than all genes
# Result: P = 0.04, FC = (33/0.2)


Everything is looking good so a common next step is to 
make a counts table of just significant genes. We can do this
by taking the entire counts table and intersecting it with only 
the gene_names or gene_ids in our dox gene list. This is also known
as indexing and something we will do A LOT of :)

# Retrieving counts data for each time point only for dox genes
```{r meging counts with our sig genes - so we have counts table for all sig genes}

# First we need to load our counts matrix
counts_matrix <- read.table("/scratch/Shares/rinnclass/MASTER_CLASS/lessons/04_RNAseq_Dox/01_Mouse_dox_wt/good_class_RNAseq/pipeline_output/star_salmon/salmon.merged.gene_counts.tsv", header=TRUE, row.names=1)

# Let's make sure our counts file is the right format
view(counts_matrix)

# Ok so we see gene_name and gene_id are same in both filtered_res_df_2 and counts_matrix
# Let's intersect these two tables !

counts_dox_genes <- counts_matrix[filtered_res_df_2$gene_id,]
unique(filtered_res_df_2$gene_id)
# Isn't that handy we can index into the counts matrix and get 
# the rows that match filtered_res_df_2$gene_id and all the columns


```
# Result now we have a counts table for heatmps of just sig genes


########## exeRcise ############
# 1) Determine if any of these properties are significantly different
# Between the dox genes and all genes:

1) padj 
2) lfcSE
3) log2foldChange

# 2 Make a results data frame of genes that are P < 0.01 & that change greater that 4 fold (up or down)
Then retreive all the counts from all the replicates for these genes
########## exeRcise ############




