---
title: "0X_counts_stuff"
author: "JR"
date: "2024-05-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Note that there is only one value for each gene but we had three replicates?
That is becuase when we use results() function we get the mean across replicates 
if we want all the values we need to use counts
```{r}
# First let's see what are column names are for each replicate:
colnames(counts(mouse_long_dox_dds, normalized = FALSE))
# They all start with WT_0

# cool now we can grep all the 0 timepoint values using "^WT_0_"
counts_0 <- counts(mouse_long_dox_dds, normalized = TRUE)[, grep("^WT_0_", colnames(counts(mouse_long_dox_dds, normalized = TRUE)))]

# Let's grab the counts for 12 hour time point as well
counts_12 <- counts(mouse_long_dox_dds, normalized = TRUE)[, grep("^WT_12_", colnames(counts(mouse_long_dox_dds, normalized = TRUE)))]

# Nice now we have an object of each replicate for 0 hour counts and 
res_counts <- results(mouse_long_dox_dds, contrast = c("time_point", "12", "0"))
significant_genes_counts <- res_counts[res_counts$padj < 0.05, ]

# we see NAs let's get rid of them
res_counts <- res_counts[!(is.na)(res_counts$padj), ]

significant_genes_counts <- res_counts[res_counts$padj < 0.05, ]



plotMA(res, main = "MA Plot: 0 vs 12 hours", ylim = c(-5, 5))
points(significant_genes$baseMean, significant_genes$log2FoldChange, col = "red")

# Cool this gave us fold change versus expression level or MA plot
# This is a standard first plot and as usual the more expressed
# the less the fold change
```

# Now let's plot the standard deviation across replicates between 0 and 12 hours
```{r}
# we are going to subset using %in% the signifcant genes from the counts object with all three replicates.
counts_0_sig <- counts_0[rownames(significant_genes_counts) %in% rownames(counts_0), ]

# Same for 12 hours
counts_12_sig <- counts_12[rownames(significant_genes_counts) %in% rownames(counts_12), ]

```

# NOw we are going to calcualte the mean and standard deviation across replicates
```{r}
# mean
mean_0 <- apply(counts_0_sig, 1, mean)

# standard error
se_0 <- apply(counts_0_sig, 1, function(x) sd(x)/sqrt(length(x)))

# same for 12 hour counts
mean_12 <- apply(counts_12_sig, 1, mean)

# standar error 12 hours
se_12 <- apply(counts_12_sig, 1, function(x) sd(x)/sqrt(length(x)))
```

# Now plotting 
again we will go into this more just to see our results for now

```{r}
# Plot 0 vs 12 hours with error bars
library(ggplot2)

df <- data.frame(
  gene = rownames(significant_genes_counts),
  time_point = c(rep("0", length(mean_0)), rep("12", length(mean_12))),
  mean = c(mean_0, mean_12),
  se = c(se_0, se_12)
)

ggplot(df, aes(x = time_point, y = mean)) +
  geom_point(aes(color = gene), position = position_dodge(0.1)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, color = gene), position = position_dodge(0.1)) +
  theme_classic() +
  labs(x = "Time Point", y = "Normalized Counts")


```


