---
title: "Deseq2_mouse_Dox_data"
author: "JR"
date: "2024-01-12"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal to get our essential files in order to run deseq2!
# Oh and to learn some basic R skills towards our goals!
The input to DESEQ2 is going to want "counts" these came from the program
Salmon from the NF_CORE pipeline output. 

# Refresher of where is my data is here:
00_Where_is_R_data.Rmd

Salmon calculates the number of reads over a gene. We also need to consider which 
version of the gene we want. Genes come in different "isoforms" which have extra or less bits
of RNA added to them. These annotations are a lot of work and difficult, but Salmon makes this easier!


# More on salmon transcript quants, this is a basic outline of how Salmon works 
Especially if you wanted to use it outside NF_CORE pipeline for RNAseq
https://combine-lab.github.io/salmon/getting_started/

For this we will use "merged_salmon_gene_counts" that gives
the representative version of a gene that is being detected as expressed in the samples (in this case mESCs exposed to dox).
This is just to note that you can also try and determine how isoforms change over time - but to get started we will keep it 
simple, thanks to the program Salmon that ran in our pipeline.

# First things first: let's get the RNAseq counts for all the genes in our GTF input file to the RNAseq pipeline.
The counts we want are called "salmon.merged.gene_counts.tsv" they are in your NF_CORE pipeline output directory:
......./pipeline_output/star_salmon/

Please locate where this directory is in your output (helpful hint: copy and paste path to location using pwd)

# Our first R function : read.table
This function in R alows us to read in a table. There are many ways to do this but we will start with the basics
Functions also have parameters to input. in this case we have a header and will call that, and where the rownames start).

Note that the <- sets a variable. 

# Let's run our first function to get our salmon mergened gene counts in a table
# 1) Read in and clean up counts table
```{r loading and cleaning up count data}
# Let's take a look at the read.table function
?read.table

# we need to make the function write to an output object. To do so we use the 
# <- assisgnement operator, here we are calling the salmon merged gene counts "counts_matrix"
# Then we will call the read.table() function that first requires a file path to the counts in quotes
# Note that in R functions follow with () where the parameter inputs are placed.
# Then we will call up different parameters with a comma, in this case header = TRUE, row.names =1
counts_matrix <- read.table("/scratch/Shares/rinnclass/MASTER_CLASS/lessons/04_RNAseq_Dox/01_Mouse_dox_wt/good_class_RNAseq/pipeline_output/star_salmon/salmon.merged.gene_counts.tsv", header=TRUE, row.names=1)

# Nice, now we have an object called counts_matrix in our Environment Panel (top right)
# Let's take a look 
View(counts_matrix)
```

# 2) Making a dataframe of gene_id and gene_name
Now let's make a table of gene_id and gene_name from the first two columns of counts_matrix
The reason being that the gene_id is a unique identifying number and the gene_symbol is more intuitive name.
We will use this a lot throughout our analyses.

Note: that the gene_id is currently in a "meta" format so we need to move it into the table.
To do this we will use the rownames function
```{r create g2s object}

# Creating a dataframe with gene_id and gene_symbol
# note creating columns with name = data
g2s <- data.frame(gene_id = rownames(counts_matrix), gene_symbol = counts_matrix[ , 1])

# Let's save this using write.csv
write.csv(g2s, "results/g2s.csv")
```


```{r loading and cleaning up count data}
# The first two columns are gene_id and gene_name
# Those will be very handy let's make an object of these

g2s <- data.frame(
  gene_id = rownames(counts_matrix),
  gene_name = counts_matrix[, 1]
)

# Note that rows or columns in black are "meta" meaning not really part of the matrix - just a label of sorts
# Note that there is a column with characters for gene name - we want a pure number matrix, but can have meta labels.
# Let's remove the gene name column for now and we will bring it back later.

# removing gene names by indexing rows and columns via object[ rows, columns ]
# here we take all the rows and remove the first column with -1 ]
counts_matrix <-counts_matrix[, -1] 
# or 
counts_matrix <- counts_matrix[, -which(names(counts_matrix) == "gene_name")]
# or
counts_matrix <- counts_matrix %>% 
  select(-gene_name)

# There are many ways to do the same thing - keeping column name is safer
# IMPORTANT: if you run the [ , -1] again it will keep removing cols

# turning into a matrix using as.matrix() function
counts_matrix <- as.matrix(counts_matrix) 
View(counts_matrix)

# Rounding the numbers with the round() function
counts_matrix_rounded <-round(counts_matrix)
View(counts_matrix_rounded)

# Note meta nature of cols and rows

```

# Nice now we have a good matrix of count values for each gene across all dox exposure times.
Next we are going to filter the counts matrix to only include genes with any expression at any time.
To do so we will learn how to index tables and use logical arguments - all in one line of code !

counts_filtered <- counts_matrix_rounded[rowSums(counts_matrix_rounded) > 1,]

# Here is what is going on in that one line of code - it's a lot, but good R skills :)
(1) create object "counts_filtered" with assignment operator <-
(2) we are going to index into "counts_matrix_rounded" using "[ ]" indexing.
This basically means we are going into that object and anything between [ ] is what we want back
(3) we call function rowSums() which sums the values across all rows (genes)
(4) rowSums needs to know what to sum on which is counts_matrix_rounded inside the rowSums() function
(5) We can add a logical argument as well. In this case any sum across a row that is greater than 1.
(6) *** indexing is [rows,columns] so far we have just selected the rows we want. 
To include all the columns we can just leave it blank after the comma.
So in the end we have [rows, and all the columns originally present]

# Let's filter to genes that have at least one value!

# 3) Filtering counts 
```{R filtering}

# Filtering so Deseq is not getting too many 0's for normalization
counts_filtered <- counts_matrix_rounded[rowSums(counts_matrix_rounded) > 1, ]

## Check out values of counts_matrix and counts_filtered in Environment panel (top right)
# How many genes were filtered out by this indexing? Hint compare counts_matrix_rounded with counts_filtered

```

# Well done we learned some Rskills while making input counts table and sample files !
Now let's save all of our work ! We have made a few key objects:
- counts filtered
- counts_matrix 
- deseq_samples

We will start our next steps toward differential gene expression analysis with DESEQ2.
So let's save them in a way we can easily reload in our next session.

# 4) Saving data in .Rdata format!
First by making a results folder
Check your working directory and set accordingly :) !

```{r saving env objects into .RData structure}

save(counts_filtered, counts_matrix, g2s, file = "results/count_files.RData")
# awesome - now we never have to load in consensus peaks again :)
```

# erase environment and reload!
```{r reload}

load("results/count_files.RData", verbose = T)
# super handy 
```

# congratulations you have made some result files using Rskills !

