---
title: "04_deseq_sample_input_file"
author: "JR"
date: "2024-04-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(tidyr)
library(dplyr)


```


# Goal: learn Rskills to create a 'colData' input file for DESEQ2
DESEQ2 not only needs the counts of RNA per gene, but also what each sample is.
DESEQ has a parameter called "colData" that tells Deseq which timepoint and replicate
is associated with which sample. 

So all we need for DESEQ2 is input counts and colData. We already got our counts in order so lets load those!
```{r loading counts objects}

load("results/count_files.RData", verbose = T)

#IMPORTANT Raw counts file
counts_matrix <- read.table("/scratch/Shares/rinnclass/MASTER_CLASS/lessons/04_RNAseq_Dox/01_Mouse_dox_wt/good_class_RNAseq/pipeline_output/star_salmon/salmon.merged.gene_counts.tsv", header=TRUE, row.names=1)


# Remove gene name column
counts_matrix <-counts_matrix %>% 
  select(-"gene_name") 
View(counts_matrix)

```


# Let's use our counts table to make the colData for DESEQ2 
We will make a dataframe that has sample_id, timepoint and replicate
A dataframe is like a excel sheet.

# 1) Create a sample dataframe
```{R colData file for DESEQ2}
# First we will create a dataframe from counts matrix that includes all the column names
# To do this we call the data.frame() function and inside make col called sample_id
# that is equal to the names of the columns in count matrix (e.g., WT_0_1)
deseq_samples <- data.frame(
  sample_id = colnames(counts_matrix))

# NOTE: this is a quick way to change cols in to rows !

# let's take a look
View(deseq_samples)

# There is good information in each sample_id and we can use code to separate out.
# First we will make an object called "split_values" to split sample_id by any underscore
# Note we are indexing into deseq_samples using '$' to denote index $column_name
# Note we use the function strsplit() to go into sample_id col and split at " _ "
split_values <- strsplit(deseq_samples$sample_id, "_")

# Let's take a look 
View(split_values)
```

# We just created a new dataframe to input our samples into DESEQ2
Note we used the sample_id to keep track of the original naming
This helps with reproducibility for tracking names acorss code.
Let's keep using this practice and get more sample info from sample_id

# 2) Learning Rfuncitons & indexing to retreive name info
```{R colData file for DESEQ2}
# Now we are going to learn another key function "sapply()"
# This is essentially a built in for loop to go over all rows and perform the same function.
# So here we will go through each row of split_values and run a "generic function(x)" 
# We will then retain the second item which is the time point value in sample_id
time_values <- sapply(split_values, function(x) x[[2]])

# Similar to above we are using sapply to grab the third fragment in split_values
# after two breaks at each "_" there are three parts - here we are grabbing the 3rd 
# NOTE indexing the function of x, which number from the split to a specific number with [[ ]]
replicate_values <- sapply(split_values, function(x) x[[3]])

# Ok now we can do more indexing and make more columns in our deseq_samples dataframe!
# The $ can allow you to index into a specific column or even create a new one.
# so we are making a new column by adding $time_point onto deseq_sample. 
# Then we fill this column by using the assignment operator to place time points in
deseq_samples$time_point <- time_values

# let's take a look:
View(deseq_samples)

# Now let's add another column for replicate
deseq_samples$replicate <- replicate_values

# let's take a look:
View(deseq_samples)

# Again note the "meta" nature of row and col names

```
# Nice we just created an index for DESEQ2 to figure out which sample is which!
The dataframe "deseq_samples" will now be used as required input into DESEQ2
DESEQ2 requires a paramater 'colData'  and we will point this to "deseq_samples"


# TIME TO FACTOR columns (IMPORTANT for DESEQ2)
So we now have all our sample information and just need to learn one more factor:
factor

# Let's factor timepoint and replicate for differential expression analysis
You can think of a "factor" as a way of putting a meta tag on data. 
It's actually required for DESEQ to do analysis only on data that has been factored!
We want to analyze time and replicates (OR MAYBE JUST TIME).

So let's make time point a factored column in deseq_samples

# 3) Factoring columns
```{R facotring timepoint and replicate column}

deseq_samples$time_point <- factor(deseq_samples$time_point)
deseq_samples$replicate <- factor(deseq_samples$replicate)

```

# Nice let's save our work
```{r saving sample sheet}

save(deseq_samples, counts_matrix, file = "results/deseq_samples.RData")


```
# Note 
We could have resaved and rewrote count_files.RData
That's not a great idea as something could have changed from the original 
file during this session. So it is best saved in the file where the code makes the object.


