---
title: "02_installing_R_packages"
author: "JR"
date: "2024-03-17"
output: html_document
---

# Here we are going to install our first packages !

There are two repositories filled with all kinds of good "software" for R. Or packages.
There are lots of powerful packages we will be using.

# CRAN
The comprehensive R archive network (CRAN) is located here:
https://cran.r-project.org/

Installing packages from here is built into R and doesn't need extra steps.
You simply run the install.packages() function from the R console.

# Tidyverse is a great package for clean, logical and readable coding.
It is located in CRAN and we will be using this form of R coding throughout this class.
We will discuss more as we start coding.

For now let's install it !
On your R console (bottom left pane in R studio) type in:
```{r}

install.packages("tidyverse")
# hint enter

```
Voila you should see tidyverse installing !
You can also install each package within tidyverse seperate.
# This is recommended for CU Students on R Studio Pro
# To this end just repeat the above install.packages for the following
(remember to use quotes around the package name.)

install.package("dplyr")
install.package("IRanges")
install.package("ggplot2")
install.package("purrr")
install.package("readr")
install.package("tibble")
install.package("tidyr")

# Bioconductor is another key repository for R packages - geared toward Bioinformatics.
For example DEseq2 is a package built to analyze RNAseq and we will be using this for class.

However Bioconductor is not built directly into R - so we need to install an installer :)

# Install BiocManager:
In your R console (" > " prompt)
```{r}

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.18")

```
Voila BiocManager is installed !

Now we will use the R command "BiocManager::install()" To install DEseq2.
Simply go to the console and type:

```{r}
BiocManager::install("DESeq2")
```
Voila DEseq2 is installed - may take 10-15 minutes.

# First let's check they are fully installed.
On the bottom right pane there is a tab called "packages" 
If you click into it - look for tidyverse and DEseq2.

Also you can go to the console and start typing "tidy" then hit tab and if tidyverse shows
up as an option it is installed ! Same for DEseq2.

As we move forwrad we will load these packages at each session (only install one time)
This will go into a special section of the .RMD files we are creating.
This section will look like this :


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(DEseq2)
library(dplyr)
# continue the list of all needed packages

```
We will run this first section and it will load the packages that have library(packagename)
# Note when loading you don't need quotes around package name but you do when installing.

Congratulations you have now installed your first R packages and two really good ones !!



